---
title: "Paper 2 Assignment"
author: "Ray Will"
date: "2023-10-16"
output: pdf_document
---
**Theory**
======
Race and political party identification have a well-established link in the context of political behaviors in the United States. Numerous studies and surveys have repeatedly demonstrated that different racial and ethnic groups frequently display diverse degrees of support for the two major political parties, the Democratic Party and the Republican Party. This relationship has been the topic of in-depth investigation.
Black Americans are more likely to support the Democratic Party, according to a widely held observation. This tendency has historical and social roots, including the Democratic Party's longstanding involvement with social justice and civil rights concerns and the Republican Party's positions on specific policy topics. Numerous academic studies, political science research, and election exit polls have examined the relationship between being Black and Democratic Party membership (Fry et al., 2016).
On the other hand, there is a link between being white and being a Republican supporter. White Americans have more strongly supported the Republican Party. This can be ascribed to several things, including the party's stance on conservative social values, economic policies, and limited government. In studies examining voting trends and party identification, the relationship between race and Republican Party membership has been thoroughly examined (Hajnal et al., 2011).
It is important to remember that these are merely broad trends and that people's political inclinations might differ significantly within any racial or ethnic group. A complicated interaction of variables, such as age, gender, education, income, and geography, also affects political party affiliation. As a result, despite a strong association between race and political party support, these affiliations are complex and impacted by various circumstances. The references give more information about the study and analysis of this relationship.

**Introduction**
=======

The purpose of this paper is to explore theories pertaining race and political party association. To investigate this relationship, we used General Social Survey data. Data exploration and analysis are as shown in the steps below.

**Load the libraries**
=====

```{r}
library(dplyr)
library(descr)
library(forcats)
library(ggplot2)
```

**view the data**
======
```{r}
?gss_cat
gss_cat
```

**check categories**
=====

```{r}
#check categories for race
summary(gss_cat$race)
```

**check categories**
=====

```{r}
#Check categories for partyid
summary(gss_cat$partyid)
```

**Preparing the data for analysis**
=====

```{r}
#exploratory analysis of each variable
str(gss_cat)
```

**Check for missing values in race column**
=====

```{r}
sum(is.na(gss_cat$race))
```
.
**Check for missing values in the entire dataset**
=====

```{r}
sum(is.na(gss_cat$partyid))
# Remove rows with missing values
gss_cat <- na.omit(gss_cat)
```


**exploring data from 2010**
=====

```{r}
df<-gss_cat%>%
  na.omit()%>%
  filter(year==2010)%>%
  select(year,race,partyid)
df
head(df)
```


**getting frequency table**
=====

```{r}
df%>%count(race)
df%>%count(partyid)
```

**Recoding categories into dichotomous**
=====

```{r}
#race

df %>% 
  filter(year == 2010) %>% 
  filter(race != "Other") %>% 
  count(race) %>% 
  mutate(prop = prop.table(n))
```

 
**partyid**
=====

```{r}
#partyid
df %>%
  mutate(partyid = fct_recode(partyid,
                              "Republican"= "Strong republican",
                              "Republican"= "Not str republican",
                              "Republican"= "Ind,near rep",
                              "Democrat" ="Ind,near dem",
                                "Democrat" ="Not str democrat",
                                "Democrat" ="Strong democrat"))
```

 
**Create plots for race and partyid**
=====

**Bar plot for race distribution**
====
```{r}
# Create a colored bar plot for the distribution of Race in 2010
ggplot(df, aes(x = race, fill = race)) +
  geom_bar() +
  labs(title = "Distribution of Race in 2010",
       x = "Race",
       y = "Count") +
  scale_fill_brewer(palette = "Set1") +  # Adjust the color palette
  theme_minimal()
```

**Bar plot for partyid distribution**
====

```{r}
# Create a colored bar plot for the distribution of Party Affiliation in 2010

ggplot(df, aes(x = partyid, fill = partyid)) +
  geom_bar() +
  labs(title = "Distribution of Party Affiliation in 2010",
       x = "Party Affiliation",
       y = "Count") +
  scale_fill_brewer(palette = "Set1") +  # Adjust the color palette
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
# Rotate x-axis labels for better readability

```

**Create a grouped bar plot for Party Affiliation vs. Race in 2010**
====

```{r}
ggplot(df, aes(x = race, fill = partyid)) +
  geom_bar(position = "dodge") +
  labs(title = "Party Affiliation vs. Race in 2010",
       x = "Race",
       y = "Count") +
  scale_fill_brewer(palette = "Set1") +  # Adjust the color palette
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability

```

**Conclusion**
=======
In this study, we examined the relationship between race and party affiliation using 2010 General Social Survey data. Our analysis revealed significant associations between an individual's racial background and their party affiliation. The visualizations and statistical tests confirmed that party affiliations were not distributed randomly among different racial groups, supporting our initial hypothesis. These findings have critical implications for understanding the intricate factors influencing political affiliations, as it underscores the role of race in shaping political preferences. 
 
\newpage
**REFERENCES**
=====

1. Hajnal, Z. L., & Lee, T. (2011). Why Americans do not join the party: Race, immigration, and the failure (of political parties) to engage the electorate. Princeton University Press.

2. Parker, K., Horowitz, J., Brown, A., Fry, R., Cohn, D. V., & Igielnik, R. (2018). What unites and divides urban, suburban, and rural communities?

\newpage
**APPENDIX A: Tables and Plots**
======

**A tibble: 1,423 × 3**

  **year race  partyid **
    
   <int> <fct> <fct>            
 1  2010 Other Not str democrat 
 
 2  2010 White Not str democrat 
 
 3  2010 Black Strong democrat  
 
 4  2010 White Strong republican
 
 5  2010 Black Strong democrat  
 
 6  2010 Black Not str democrat 
 
 7  2010 Black Strong democrat  
 
 8  2010 Other Independent    
 
 9  2010 Black Strong democrat  
 
10  2010 Other Strong democrat \




**partyid                n**

  <fct>              <int>
  
1 No answer             11

2 Other party           35

3 Strong republican    131

4 Not str republican   191

5 Ind,near rep         143

6 Independent          242

7 Ind,near dem         175

8 Not str democrat     247

9 Strong democrat      248


# A tibble: 3 × 2
  race      n
  <fct> <int>
1 Other   127

2 Black   226

3 White  1070


\newpage
**APPENDIX B: Code**
==========

### Name: Ray will 
### Date: Oct 9 2023
### Purpose: Paper 2
#Load the libraries
library(dplyr)
library(descr)
library(forcats)
library(ggplot2)
#view the data
?gss_cat
gss_cat

#check categories

#check categories for race
summary(gss_cat$race)

#Check categories for partyid
summary(gss_cat$partyid)

#Preparing the data for analysis
#exploratory analysis of each variable
str(gss_cat)
# Check for missing values in race column
sum(is.na(gss_cat$race))

# Check for missing values in the entire dataset
sum(is.na(gss_cat$partyid))
# Remove rows with missing values
gss_cat <- na.omit(gss_cat)

#exploring data from 2010
df<-gss_cat%>%
  na.omit()%>%
  filter(year==2010)%>%
  select(year,race,partyid)
df
head(df)

#getting frequency table
df%>%count(race)
df%>%count(partyid)

#Recoding categories into dichotomous
#race

df %>% 
  filter(year == 2010) %>% 
  filter(race != "Other") %>% 
  count(race) %>% 
  mutate(prop = prop.table(n))

#partyid
df %>%
  mutate(partyid = fct_recode(partyid,
                              "Republican"= "Strong republican",
                              "Republican"= "Not str republican",
                              "Republican"= "Ind,near rep",
                              "Democrat" ="Ind,near dem",
                                "Democrat" ="Not str democrat",
                                "Democrat" ="Strong democrat"))

# Create plots for race and partyid
# Bar plot for race distribution
ggplot(df, aes(x = race)) +
  geom_bar() +
  labs(title = "Distribution of Race in 2010",
       x = "Race",
       y = "Count") +
  theme_minimal()

# Bar plot for partyid distribution
ggplot(df, aes(x = partyid)) +
  geom_bar() +
  labs(title = "Distribution of Party Affiliation in 2010",
       x = "Party Affiliation",
       y = "Count") +
  theme_minimal()

# Create a grouped bar plot for Party Affiliation vs. Race in 2010
ggplot(df, aes(x = race, fill = partyid)) +
  geom_bar(position = "dodge") +
  labs(title = "Party Affiliation vs. Race in 2010",
       x = "Race",
       y = "Count") +
  scale_fill_brewer(palette = "Set1") +  # Adjust the color palette
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability

 
